\documentclass[dvipsnames,10pt,aspectratio=169]{beamer}
\usetheme{metropolis}           % Use metropolis theme


\usepackage{xcolor}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{fontspec}
\usepackage{proof}
\usepackage{tikz-cd}
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{hyperref}
\hypersetup{colorlinks,linkcolor=blue,urlcolor=blue}
\setmonofont[Scale=0.7]{DejaVu Sans Mono}

\bibliographystyle{alpha}
\setbeamerfont{bibliography item}{size=\footnotesize}
\setbeamerfont{bibliography entry author}{size=\footnotesize}
\setbeamerfont{bibliography entry title}{size=\footnotesize}
\setbeamerfont{bibliography entry location}{size=\footnotesize}
\setbeamerfont{bibliography entry note}{size=\footnotesize}
\setbeamertemplate{bibliography item}{}


%% --------------------------------------------------------------------------------

\title{Universes In Type Theory}
\date{January 29, 2021, PLC Department Workshop}
\author{\vspace{-1em}András Kovács\footnote{The author was supported by the European Union, co-financed by the
European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).\vspace{0.5em}}}
\institute{Eötvös Loránd University, Budapest}
\begin{document}
\maketitle

%% ------------------------------------------------------------

\begin{frame}{Russell's paradox, or size issues}

There is no set $S$ in a consistent set theory such that $x \in S$ iff $x \notin x$.
\vspace{1em}

Alternate formulations:
\begin{itemize}
  \item Barber's paradox
  \item No set of all sets
\end{itemize}
\vspace{1em}

Related:
\begin{itemize}
  \item Real numbers are uncountable
  \item Gödel's first incompleteness theorem
  \item Undecidability of halting
\end{itemize}
\vspace{1em}

{\footnotesize (All instances of \href{https://arxiv.org/pdf/math/0305282.pdf}{Lawvere's fixed point theorem)}}
\end{frame}

%% ------------------------------------------------------------

\begin{frame}{Solution with universes}

Solving Russell's paradox: set comprehension can only define a \textbf{subset} of a set.
\vspace{1em}

But sometimes we still want to quantify over \textbf{all sets} of
Zermelo-Fraenkel set theory.
\vspace{1em}

We can assume the existence of a very large set such that all of ZF can fit there,
which is called a \textbf{universe}. Then, just quantify over elements of the
universe. The resulting theory is stronger than ZF.
\vspace{1em}

If we want quantify over sets which \textbf{don't} fit (or universes
themselves), we can just assume even more universes.

\end{frame}


\begin{frame}[fragile]{Universes in type theory}

Agda:
\begin{verbatim}
  Bool          : Set₀
  Set₀          : Set₁
  Set₁          : Set₂
  (Bool → Set₀) : Set₁
  (Set₀ → Bool) : Set₁
\end{verbatim}

Assuming \texttt{Set₀ : Set₀} implies contradiction (by Russell-like argument).
\vspace{1em}

{\small(But type theory with \texttt{Set₀ : Set₀} is still a perfectly good Turing-complete
programming language)}
\vspace{1em}

\end{frame}

\begin{frame}[fragile]{Universes in type theory}

The basic system can be tedious:
\begin{verbatim}
  id₀ : (A : Set₀) → A → A
  id₁ : (A : Set₁) → A → A
  ...
\end{verbatim}

In Coq/Agda/Lean, various extra features are used.
Example: universe polymorphism in Agda:
\begin{verbatim}
  id : (l : Level)(A : Set l) → A → A
  id l A x = x
\end{verbatim}


\end{frame}

\begin{frame}[fragile]{Design choices and variations}

\textbf{How many universes?} Agda/Coq: countably many.
\vspace{1em}

\textbf{Are universes totally ordered?} Agda/Coq: yes.
\vspace{1em}

\textbf{What kind of level polymorphism?}. Coq: bounded polymorphism. Agda: no
bounds allowed. Bounded example:
\begin{verbatim}
  myId : (l : Level) → l < 3 → (A : Set l) → A → A
  myId l p A x = x
\end{verbatim}

\textbf{What kind of operations are available on levels?} Agda is more liberal than Coq. Example:
\begin{verbatim}
  ℕtoLevel : ℕ → Level
\end{verbatim}


\textbf{Are universes cumulative} Agda: no. Coq: yes. Cumulativity: whenever \texttt{A : Set i}, we also have \texttt{A : Set (i + 1)}.

\end{frame}

\begin{frame}{Research goals}

We want to know that each point in the design space makes sense.
\vspace{1em}

Making sense:
\begin{itemize}
  \item Logical consistency.
  \item Is the type theory a proper programming language? Programs should compute to values
        and not get randomly stuck.
  \item Is type checking decidable? (I don't plan to cover this in research).
\end{itemize}

Approach: use generic framework to cover as many features/variations as
possible. Prove that everything in the framework is sensible.


\end{frame}


\begin{frame}{Results}

Framework covers:
\begin{enumerate}
  \item Universe levels may come from any well-ordered set.
  \item Polymorphism with bounds is allowed.
  \item Universes are cumulative.
  \item Levels can be manipulated arbitrarily as program data (WIP).
\end{enumerate}

Things we need to assume in order to prove the framework sensible:
\begin{enumerate}
  \item \textbf{Two} universes.
  \item Types of certain infinitely branching trees (W-types).
\end{enumerate}

Potential further applications (in future work):
\begin{itemize}
  \item Information flow type systems (``secure'' and ``public'' levels).
  \item Staged compilation (``runtime'' and ``compile time'' levels).
\end{itemize}

\end{frame}

\begin{frame}

This talk is a side project intended for FSCD 2021 submission (in February).
\vspace{1em}

Other 2020 publications:
\begin{itemize}
  \item LICS 2020, with Ambrus: ``Large and Infinitary Quotient Inductive-Inductive Types''
  \item ICFP 2020: ``Elaboration with First-Class Implicit Function Types''
  \item TYPES 2019 post-proceedings, with Ambrus and Ambroise Lafont: ``For
    Finitary Induction-Induction, Induction is Enough''
\end{itemize}

\end{frame}


\begin{frame}

\begin{center} {\large Thank you!} \end{center}

\end{frame}


%% \begin{frame}[allowframebreaks]{References}
%%   \bibliography{references}
%% \end{frame}

\end{document}
